A Component-Based Sensor Network for \\ Environmental Monitoring

### 2009-03-13 docs/icccn2009/paper.tex
This is my paper revision in all sections. Please, diff this revision with the previous one for better capturing the details.
git-svn-id: svn://localhost/netbeams/trunk@385 559edde2-a4d4-44e0-96fc-4035e40d7396

A wireless sensor network consists of spatially distributed autonomous
devices using sensors to cooperatively monitor physical or
environmental conditions at different locations \cite{roemer:2004}.
Much research effort is focused on building wireless ad-hoc networks
where each sensor node participates in a multi-hop routing algorithm.
One the earliest examples of a wireless sensor network is the
SmartDust project \cite{smartdust:2001} where one objective was to
create autonomous sensing and communication in a cubic
millimeter-sized device.  While most wireless sensor networks focus on
small sensing devices such as the SmartDust mote, there exist a wide
variety of off-the-shelf commercially available sensors that are used
by environmental researchers. In contrast to the small sensing
devices, sensors for environmental monitoring are typically on the
order of one to two feet in length weighing several pounds. Our
focus is on building an end-to-end infrastructure for environmental
monitoring using specialized sensing devices.

Environmental monitoring depends on the reliable collection of
measurements from remotely deployed sensors and the rapid transfer of
those measurements to data centers. Typically, sensors are deployed in
out-of-the-way locations, where physical access is limited and
connections for telemetry are poor or nonexistent. Maintaining the
data stream from these sensors places demands on human resources,
requiring site visits to service the sensors and to download data
stored on internal memory.  Consequently, there is a lag of days to
months between the time the sensors perform the measurements and the
time the measurements become available. Advances in battery life and
anti-fouling technology have extended service cycles for field
sensors, with the unwanted result of further delaying access to
monitoring data.  In addition, the dependence on field site visits to
alter of sensor characteristics, such as sampling rate, prevents
adjustment in monitoring strategy in response to rapidly developing
events (e.g. oil spills, harmful algal blooms).

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides means to program
and interrogate sensors at the remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
real-time. In the following sections we give an overview of the NetBEAMS
architecture and implementation.  Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network.
Section \ref{SEC_NETBEAMS} describes NetBEAMS; a practical application
of the DSP. Section \ref{SEC_CONCLUSION} concludes this paper.

@@

Each probe may have multiple sensors.  A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode.  Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode the Sonde is usually connected via 650
MDS Display/Logger or a serial cable connected to a PC.  The sampling
frequency is usually set to a high frequency in this mode.  Unattended
mode is performed usually when the Sonde is deployed offshore or in a
remote body of water.  The sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). The Sonde is configured via a
set of menus that can be accessed through a terminal session over
RS232. Through these menus, the Sonde can be configured to start
logging data to an internal file.  To view the data as it's being
captured, the option to show live data is chosen and the data can then
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
access to their measurements.  Typically, the sensors are autonomous
in the sense that they can run for a certain time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but also results are in a serious
time lag between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurement. We identify the following requirements that an
end-to-end infrastructure supporting devices such as the YSI should
offer:

\begin{itemize}
\item Allow near-realtime access to sensor measurements.
\item Allow remote servicing that include changing of configuration
  parameters as well as software upgrades.
\item Support cost-effective long distance communication link.
\item Can be deployed in remote locations in unprotected outdoor
  conditions.
\end{itemize}

In the next section we describe a software architecture that provides
a foundation to fulfill these requirements.

@@

Sensor networks, especially of environmental sensors, are normally
deployed over a geographically wide area where the communication
infrastructure is not always reliable or, at least, not countinuously
available. As the sensor network grows in size and complexity
scalability, reliabilty and maintainability become critical issues.
Additional components might be added to the network, sensor
configuration might be altered, or new versions of components might be
deployed. The sensor network must be able to support such adjustments
without significant overhead for maintainers.  From system
development and integration point of view, interfaces among components
and between components and the network should be simple. Complex
interaction can be constructed on top of these simple interfaces.  In
the following we discuss how our proposed architechure addresses those
issues.

@@

Our architecture is based on a micro-kernel approach. The core
functionality is limited to a minimum and all extra services are built
as plug-and-play modules on top of the micro-kernel. With the help of
this approach we can keep the basic infrastructure compact while
allowing customization via special-purpose modules. In our terminology
we call the basic foundation the \emph{Data Sensor Platform}
(DSP). The scope of the DSP is one address space of an execution
platform. A sensor network is built from several DSPs that are linked
with each other.

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication paradigm within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding simple but powerful communication
instructure.  We distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have. It is possible for a DSP
Component to act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules. Rules can be configured statically
or dynamically. Rules for routing messages can be based on different
information: producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a module that interfaces with a physical sensor. Such
a module reads data from the sensor and converts the data into a
message that is forwarded to the DSP. Because this module generates
data with respect to the DSP, it is a data producer. An example for a
data consumer is a converter module that converts the internal,
DSP-specific data format to a format suitable for consumption by
external applications. One such common format used in the domain of
environmental monitoring is OpenDAP \cite{opendap01}.  Since the
converter module accepts messages from the DSP, it acts in the role of
a data consumer. Finally, a DSP Component implementing persistency is
an example of a module that acts in the roles of both data producer
and consumer. It accepts messages from the DSP and stores them in a
database. At a later time, it may act as a data producer by resending
previously stored messages.

@@

By implementing the above Java interface, the resuling code
effectively becomes a DSP Component that can be deployed on a DSP at
runtime. A component is identified by its type (line 3). Only one
component of a certain type per DSP is permissible. A type identifier
must be unique across all DSP Components in a DSP.  Upon deploying a
component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 5) and stopped (line
6) multiple time. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (line
7). The message structure will be explained in the following section.
One important feature of the DSP is the ability to redeploy (maybe
with a new version) at runtime. This allows DSPs to be updated
remotely. This process is managed through messages sent to a special
DSP Component that is able to manage its local DSP.

@@

Message passing is the main paradigm for exchanging information within
a DSP. After producing data, a DP wraps that data in the body
of a message, so that DC components can receive the message and
consume the data contained in the message's body. When it comes to passing
messages to a remote DSP instance, messages are serialized in XML
\cite{xml2000} and automatically deserialized for DC located in the
remote DSP. In order to serialize and deserialize DSP messages, the
DSP infrastructure uses the XML Schema standard \cite{xml-schema2004}
to generate and validate the DSP message. Figure \ref{FIG_DSP_MESSAGE}
defines the structure of a general DSP message.

@@

The basic message structure is defined by an \texttt{AbstractMessage}
type, whose structure serves as a base schema for all different types
of messages. Any message will contain two basic attributes:
\texttt{messageID} and \texttt{ContentType}.  The former uniquely
identifies a message and the latter the data in the body of the
message. Complementary identification information such as the message
producer and consumer are located in the header of the message, which
are defined by the schema type \texttt{ComponenIdentifier}. A
\texttt{ComponentIdentifier} carries out information about the
component name and its physical host address.  Finally, the body of
the message represents the payload and may contain any data type
defined by the component developer.

Although XML schema is the de-facto standard to bind and validate XML
data to its type, the architectural infrastructure of DSP uses an XML
data binding technology \cite{xml-dbind} called JAXB \cite{xml-jaxb}.
It is used to generate a type safe API for handling in-memory object
graphs from the XML instances of DSP Messages and the content of its
data payload. Moreover, taking advantage of the data type polymorphism
of XML schema data types -- a concept borrowed from object oriented
principles -- we defined different categories of messages in order to
reflect the nature of the message content. Among those categories are
query and update messages for configuring various properties supported
by DSP Components and measurement messages that actually contain
sensor readings.

The payload of a message is itself described via an XML schema in
order to integrate it seemingly with the general message structure. A
developer of a DSP Component can define a schema that represents the
type of payload processed by that component. An example illustrating a
DSP message is given when we discuss the YSI Sonde Component.

@@

In order to keep the API symmetrical, the DSP is itself implemented as
a DSP Component, the so-called DSP Broker. While technically the DSP
Broker has the same API as any component, it serves a special role and
cannot be removed. When a DP sends a message, it actually
invokes the \texttt{deliver()} method of the DSP Broker. The broker is
responsible for routing the message to one or several DCs. It
distinguishes between local and remote destinations:

\begin{itemize}
\item Local delivery of messages: it uses local object reference to
  deliver the DSP message.
\item Remote delivery of messages: it uses a DSP Component marked as a
  default gateway to handle remote message delivery.
\end{itemize}

The selection of the DCs for a given message to where the broker will
deliver messages to is determined by a matcher.  The matcher is
responsible to find the possible set of DCs for a given DSP message,
and it relies on matching rules defined statically or at runtim. In
this way, a set of matching rules is defined in XML and stored at the
DSP run-time directory, where each matching rule is uniquely
identified by a rule ID and determines the following (see Figure
\ref{FIG_MATCHER_RULE}):

@@

OSGi \cite{osgi} is a generic platform for "dynamic module system for Java\texttrademark".
\cite{Mueller07} used OSGi as platform their sensor network.
In \cite{Baude07} OSGi is used to support large scale deployment. OSGi is flexible
enough to be used in mobile devices and in application servers. One aspect not addressed
by OSGi is distributed communication (although some extensions have
been proposed,e.g., \cite{Rellermeyer}). DSP was designed to handle remote communication and,
therefore, can complement OSGi capability. DSP Components are OSGi bundles (in fact, DSP itself
is a bundle) and can be remotely and dynamically maintained. OSGi capabilities along with DSP
design make well suitable for our project.

@@

In the following we present a practical application of the DSP
introduced in the previous section. Our environmental sensor network
is dubbed NetBEAMS (Networked Bay Environmental Assessment Monitoring
System) and allows access to sensor equipment deployed throughout the
San Francisco Bay Area. This end-to-end system is a ready-to-use
turnkey solution that is adaptable to a wide variety of sensor types
and requires no programming expertise to operate. Configuration of the
system will be possible when the sensor is deployed in the field via a
web-based interface, eliminating the need for extra site visits by
technicians between service cycles to change systems functionalities
(e.g. sampling rate).  We present the NetBEAMS infrastructure by
discussing its hardware and software components.

@@

NetBEAMS supports an environmental sensor such as the YSI Sonde,
introduced in Section \ref{SEC_BACKGROUND}. In order to access its
sensor readings, we collocate it with an off-the-shelf embedded
computing platform called Gumstix \cite{gumstix01} (far left in Figure
\ref{FIG_NETBEAMS}).  The Gumstix is an ARM architecture and comes in
various hardware configurations.  A typical Gumstix configuration
consists of a motherboard and one or more expansion boards which
connect to the motherboard via on-board buses.  The motherboards draw
less than 250 mA @4V at 400 MHz and less than 50 mA while idling,
waiting for input.

The Gumstix runs Linux 2.6 with the BusyBox utilities, and uses the
OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture.  We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP. The Gumstix
communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable since both the Sonde and the Gumstix are DTE devices.
%The serial port configuration is:

@@

We make use of the cellular phone network to establish a communication
link with the backend server. The cellular phone network has the
advantages of reaching well beyond telemetry options such as wireless
networks and being a significantly lower cost solution than satellite
communications. We make use of the Huawei E220. The E220 is a USB
modem that supports HSDPA, UMTS, and EDGE packet data services at
maximum transmission rates of 3.6Mbps, 384kbps, and 236.8kbps
respectively.  HSDPA and UMTS operate at 2100MHz while GSM, GPRS, and
EDGE operate at 900, 1800, and 1900MHz. The modem connects to the
Gumstix motherboard via a mini USB interface (supporting USB 2.0).The
Gumstix communicates with the modem using the PPP protocol and
accesses the cellular network using a data plan from AT\&T.

@@

The DSP is implemented in Java using the OSGi platform. OSGi
\cite{osgi} is a Java-based component system. So-called bundles, which
are self-contained Java JAR files, can be deployed and undeployed at
runtime via special purpose class loaders. In fact, the API of the DSP
is inspired by OSGi. All DSP Components depicted in Figure
\ref{FIG_NETBEAMS} are implemented as OSGi bundles. The DSP Broker
discussed in Section \ref{DSP_BROKER} is mapped to an OSGi bundle as well.

Our implementation of the DSP is based on the Open Source
implementation of OSGi called Knopflerfish \cite{knopflerfish01}.
Since Sun Microsystems does not support its JDK for the Gumstix, we
use JamVM \cite{jamvm01} as the Java virtual machine implementation
and GNU Classpath \cite{classpath01} for the runtime libraries. In the
following sections we provide further details of some of the DSP
Components that we have implemented as part of the NetBEAMS project.

@@

On the far left of Figure \ref{FIG_NETBEAMS} is the actual sensor; in
our case the aforementioned YSI Sonde. The YSI Sonde DP \ding{192}
uses the RS232 standard to read the physical measurements from the
device over a serial port.  This component converts the physical
measurements into a DSP-specific message. Figure
\ref{FIG_DSP_YSI_DATA} depicts the YSI Sonde's specific message format
which includes each measurement data point, described in Table
\ref{TAB_SONDE_MEASUREMENTS}.

@@

The \texttt{SondeDataContainerType} defines the payload of messages generated
by the YSI Sonde DP. As all DSP messages, it is derived from the
\texttt{AbstractMessageContent}. The payload mainly consists of the
\texttt{SondeDataType} that defines all the sensor data that the YSI Sonde is
capable of producing. The YSI Sonde DP uses the RS232 to query the
physical device for its data in regular intervals. The measurements
are then converted into a DSP message and forwarded to the DSP Broker
from there they are forwarded to the remote communication component
described in the next section.  The YSI Sonde DP can also be
configured via special DSP messages, to direct the Sonde to change its
sampling frequency.  These property messages originate from a web
management interface (component \ding{197} in Figure \ref{FIG_NETBEAMS}).

@@

All messages sent by the YSI Sonde DP need to be uploaded to a remote
database. Since the DSP itself can only handle communication within
its local DPs and DCs, we have created a matching pair of DC and DP
that can handle remote communication (components \ding{193} and \ding{194} in Figure
\ref{FIG_NETBEAMS}). As explained earlier, we make use of the cellular
phone network to upload sensor data to a remote backend server. While
the cellular phone network has sufficient coverage for more
environmental sensor networks, it is not cost effective to keep a data
link permanently open. The remote communiction components can be
configured to upload buffered sensor data only in certain time
intervals. Messages are transported via HTTP \cite{RFC2068}. The
matching DP/DC pair for remote communication has the following
properties:

\begin{itemize}
\item DSP Wire Transport Client: takes a collection of DSP messages
  and transmit them via HTTP. As a non-functional requirement, the
  interval in which this component will transmit messages to other
  components can be configured via the management interface.
\item DSP Wire Transport Server: is a HTTP server that receives the
  collection of DSP messages sent by the client, and piggy backs any
  messages destined for the remote client that may have queued up in
  the meantime.
\end{itemize}

The HTTP request and response are used to upload and download any
messages that have been queued at the respective end of the
communication link. The client and server component are capable of
buffering messages until they are sent to the remote destination
during the next communication cycle. Once messages have been
transported to their final destinations, they are unmarshalled and
passed on to the local DSP Broker for further dispatching. In this
regard the client and server are completely symmetrical. Messages are
dispatched with the help of the matcher to their final destinations.

>>>>

A wireless sensor network consists of spatially distributed autonomous
devices using sensors to cooperatively monitor physical or
environmental conditions at different locations \cite{roemer:2004}.
Much research effort is focused on building wireless ad-hoc networks, 
where each sensor node participates in a multi-hop routing algorithm.
One of the earliest examples of a wireless sensor network is the
SmartDust project \cite{smartdust:2001}, where one objectives was to
create autonomous sensing and communication in a cubic
millimeter-sized device.  While most wireless sensor networks focus on
small sensing devices such as the SmartDust mote, there exist a wide
variety of off-the-shelf commercially available sensors that are used
by environmental researchers. In contrast to the small sensing
devices, sensors for environmental monitoring are typically on the
order of one to two feet in length, weighing several pounds. Our
focus is on building an end-to-end infrastructure for environmental
monitoring using specialized sensing devices.

Environmental monitoring depends on the reliable collection of
measurements from remotely deployed sensors and the rapid transfer of
those measurements to data centers. Typically, sensors are deployed in
out-of-the-way locations, where physical access is limited and
connections for telemetry are poor or nonexistent. Maintaining the
data stream from these sensors places demands on human resources,
requiring site visits to service the sensors and to download data
stored on internal memory.  Consequently, there is a lag of days to
months between the time the sensors perform the measurements and the
time the measurements become available. Advances in battery life and
anti-fouling technology have extended service cycles for field
sensors, with the unwanted result of further delaying access to
monitoring data.  In addition, the dependence on field site visits to
alter of sensor characteristics, such as sampling rate, prevents
adjustment in monitoring strategy in response to rapidly developing
events (e.g. oil spills, harmful algal blooms, among others).

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides means to program
and interrogate sensors at the remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
real-time. In the following sections we give an overview of the NetBEAMS
architecture and implementation.  Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network while
Section \ref{SEC_NETBEAMS} describes NetBEAMS. Finally, Section 
\ref{SEC_CONCLUSION} presents our conclusions and future works.

@@

Each probe may have multiple sensors.  A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode.  Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode, the Sonde is usually connected via 650
MDS Display/Logger or a serial cable connected to a PC, where the sampling
frequency is usually set to a high frequency. In contrast, Unattended
mode is performed usually when the Sonde is deployed offshore or in a
remote body of water, where the sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). In order to be configured,
the Sonde offers a set of menus that can be accessed through a terminal session
over RS232. In this way, the Sonde can be configured to
start logging data to an internal file. In order to view the data as it's
being captured, the option to show live data is chosen and the data can then be 
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
the access to their measurements.  Typically, the sensors are autonomous
in the sense that they can run for a certain amount of time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but also results are in a serious
time lag between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurement. We identify the following requirements that an
end-to-end infrastructure supporting devices such as the YSI should
offer:

\begin{itemize}
\item Allow near-realtime access to sensor measurements;
\item Allow remote servicing that include changing of configuration
  parameters as well as software upgrades;
\item Support cost-effective long distance communication link;
\item Can be deployed in remote locations in unprotected outdoor
  conditions.
\end{itemize}

In the next section, we describe a software architecture that provides
a foundation to fulfill these requirements.

@@

Sensor networks, especially of environmental sensors, are normally
deployed over a geographically wide area, where the communication
infrastructure is not always reliable or, at least, not coutinuously
available. As the sensor network grows in size and complexity, 
scalability, reliability and maintainability become critical issues.
For instance, additional components might be added to the network, sensor
configuration might be altered, or new versions of components might have to be
deployed. As a consequence, the sensor network must be able to support such
adjustments without significant overhead for maintainers.  From system
development and integration point of view, interfaces among components
and between components and the network should be simple. As a result, complex
interaction can be constructed on top of these simple interfaces.  In
the following subsections, we discuss how our proposed architecture addresses
those issues.

@@

NetBEAMS's architecture is based on a micro-kernel approach. The core
functionality is limited to a minimum and all extra services are built
as plug-and-play modules on top of the micro-kernel. With the help of
this approach, we can keep the basic infrastructure compact while
allowing customization via special-purpose modules. In our terminology,
we call the basic foundation the \emph{Data Sensor Platform}
(DSP). The scope of the DSP is one address space of an execution
platform and a sensor network is built from several DSPs that are linked
with each other.

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication unit within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding simple but powerful communication
structure. Moreover, we distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have, as it may also 
act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules, which can be configured statically
or dynamically. Rules for routing messages can be based on different
information such as producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a module that interfaces with a physical sensor. Such
a module reads data from the sensor and converts the data into a
message that is forwarded to the DSP. Because this module generates
data with respect to the DSP, it is a data producer. An example for a
data consumer is a converter module that converts the internal,
DSP-specific data format to a format suitable for consumption by
external applications. One such common format used in the domain of
environmental monitoring is OpenDAP \cite{opendap01}.  Since the
converter module accepts messages from the DSP, it acts in the role of
a data consumer. Finally, a DSP Component implementing persistence is
an example of a module that acts in the roles of both data producer
and consumer, since it accepts messages from the DSP and stores them in a
database. Furthermore, it may act as a data producer by resending previously
stored messages at a later time.

@@

By implementing the above Java interface, the resulting code
effectively becomes a DSP Component that can be deployed on a DSP at
runtime. A component is identified by its type (line 3). Only one
component of a certain type per DSP is permissible. A type identifier
must be unique across all DSP Components in a DSP.  Upon deploying a
component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 5) and stopped (line
6) multiple times. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (line
7). The message structure will be explained in the next section.
One important feature of the DSP is the ability to redeploy itself (maybe
with a new version with updates) at runtime. This allows DSPs to be updated
remotely. This process is managed through messages sent to a special
DSP Component that is able to manage its local DSP.

@@

Message passing is the main paradigm for exchanging information within
a DSP. After producing data, a DP wraps that data in the body
of a message, so that DC components can receive the message and
consume the data contained in the message's body. When it comes to passing
messages to a remote DSP instance, messages are serialized in XML
\cite{xml2000} and automatically deserialized for DCs located in the
remote DSP. In order to serialize and deserialize DSP messages, the
DSP infrastructure uses the XML Schema standard \cite{xml-schema2004}
to generate and validate the DSP message. Figure \ref{FIG_DSP_MESSAGE}
defines the structure of a general DSP message.

@@

The basic message structure is defined by an \texttt{AbstractMessage}
type, whose structure serves as a base schema for all different types
of messages. Any message will contain two basic attributes:
\texttt{messageID} and \texttt{ContentType}.  The former uniquely
identifies a message and the latter the data in the body of the
message. Complementary identification information, such as the message
producer and consumer, is located in the header of the message and 
is defined by the schema type \texttt{ComponenIdentifier}. A
\texttt{ComponentIdentifier} carries out information about the
component name and its physical host address.  Finally, the body of
the message represents the payload and may contain any data type
defined by the component developer.

Although XML schema is the de-facto standard to bind and validate XML
data to its type, the architectural infrastructure of DSP uses an XML
data binding technology \cite{xml-dbind} called JAXB \cite{xml-jaxb}.
It is used to generate a type safe API for handling in-memory object
graphs from the XML instances of DSP Messages and the content of its
data payload. Moreover, taking advantage of the data type polymorphism
of XML schema data types -- a concept borrowed from object oriented
principles -- we defined different categories of messages in order to
reflect the nature of the message content. Among those categories are
query and update messages, where the former is used for configuring various
properties supported by DSP Components, and the latter is used for measurement
messages that actually contain sensor readings.

The payload of a message is itself described via an XML schema in
order to integrate it seemingly with the general message structure. By this
means, a developer of a DSP Component can define a schema that represents the
type of payload processed by that component. An example illustrating a DSP
message is given when we discuss the YSI Sonde Component.

@@

In order to keep the API symmetrical, the DSP is itself implemented as
a DSP Component, the so-called DSP Broker. While technically the DSP
Broker has the same API as any component, it serves a special role and
cannot be removed. When a DP sends a message, it actually
invokes the \texttt{deliver()} method of the DSP Broker. The broker is
responsible for routing the message to one or several DCs. Furthermore, it
distinguishes between local and remote destinations:

\begin{itemize}
\item Local delivery of messages: it uses local object reference to
  deliver the DSP message;
\item Remote delivery of messages: it uses a DSP Component marked as a
  default gateway to handle remote message delivery.
\end{itemize}

The selection of the DCs for a given message to where the broker will
deliver messages to is determined by a matcher.  The matcher is
responsible to find the possible set of DCs for a given DSP message,
and it relies on matching rules defined statically or at runtime. In
this way, a set of matching rules is defined in XML and stored at the
DSP run-time directory, where each matching rule is uniquely
identified by a rule ID and determines the following (see Figure
\ref{FIG_MATCHER_RULE}):

@@

OSGi \cite{osgi} is a generic platform for "dynamic module system for Java\texttrademark".
Many researchers have been using OSGi to support different requirements similar
to the DSP. For instance, \cite{Mueller07} used OSGi as a platform to their
sensor network, while \cite{Baude07} used OSGi to support large scale
deployment. As many have researchers have noticed, OSGi is flexible enough to be
used in memory-limited mobile devices or in larger application servers.
However, as being primarily focused on providing an infrastructure for
loosely-coupled module interaction, OSGi does not address distributed
communication although some extensions have been proposed, e.g.,
\cite{Rellermeyer}). As one of the contributions of this project, DSP was
designed to handle remote communication and, therefore, can complement OSGi
capability. Therefore, DSP Components are OSGi bundles (in fact, DSP itself is a
bundle) and they can be remotely and dynamically maintained, so that OSGi
capabilities along with DSP design make well suitable for our project.

@@

In the following subsection, we present a practical application of the DSP
introduced in the previous section. Our environmental sensor network
is dubbed NetBEAMS (Networked Bay Environmental Assessment Monitoring
System) and allows access to sensor equipment deployed throughout the
San Francisco Bay Area. This end-to-end system is a ready-to-use
turnkey solution that is adaptable to a wide variety of sensor types
and requires no programming expertise to operate. Configuration of the
system will be possible when the sensor is deployed in the field via a
web-based interface, eliminating the need for extra site visits by
technicians between service cycles to change systems functionalities
such as sampling rate.  We present the NetBEAMS infrastructure by
discussing its hardware and software components.

@@

NetBEAMS supports an environmental sensor such as the YSI Sonde,
introduced in Section \ref{SEC_BACKGROUND}. In order to access its
sensor readings, we collocate it with an off-the-shelf embedded
computing platform called Gumstix \cite{gumstix01} (far left in Figure
\ref{FIG_NETBEAMS}).  The Gumstix is an ARM architecture hardware and comes in
various configurations.  A typical Gumstix configuration
consists of a motherboard and one or more expansion boards that
connect to the motherboard via on-board buses.  The motherboards draw
less than 250 mA @4V at 400 MHz and less than 50 mA while idling,
waiting for input.

The Gumstix runs Linux 2.6 with the BusyBox utilities, and uses the
OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture.  We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP. The Gumstix
communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable, since both the Sonde and the Gumstix are DTE devices.
%The serial port configuration is:

@@

We make use of the cellular phone network to establish a communication
link with the backend server. The cellular phone network has the
advantages of reaching well beyond telemetry options such as wireless
networks and being a significantly lower cost solution than satellite
communications. We make use of the Huawei E220, which is a USB
modem that supports HSDPA, UMTS, and EDGE packet data services at
maximum transmission rates of 3.6Mbps, 384kbps, and 236.8kbps
respectively.  HSDPA and UMTS operate at 2100MHz while GSM, GPRS, and
EDGE operate at 900, 1800, and 1900MHz. The modem connects to the
Gumstix motherboard via a mini USB interface (supporting USB 2.0). In this
way, the Gumstix communicates with the modem using the PPP protocol and
accesses the cellular network using a data plan from AT\&T.

@@

As mentioned earlier, the DSP is implemented in Java using the OSGi
platform. The so-called bundles, which are self-contained Java JAR files, can
be deployed and undeployed at runtime via special purpose class loaders. In 
fact, the API of the DSP is inspired by OSGi. All DSP Components depicted in Figure
\ref{FIG_NETBEAMS} are implemented as OSGi bundles. The DSP Broker discussed in
Section \ref{SEC_DSP_BROKER} is mapped to an OSGi bundle as well.

Our implementation of the DSP is based on the Open-Source
implementation of OSGi called Knopflerfish \cite{knopflerfish01}.
Since Sun Microsystems does not support its JDK for the Gumstix, we
use JamVM \cite{jamvm01} as the Java virtual machine implementation
and the GNU Classpath \cite{classpath01} project for the runtime libraries. In
the following sections, we provide further details of some of the DSP
Components that we have implemented as part of the NetBEAMS project.

@@

On the far left of Figure \ref{FIG_NETBEAMS} is the actual sensor; in
our case, the aforementioned YSI Sonde. The YSI Sonde DP \ding{192}
uses the RS232 standard to read the physical measurements from the
device over a serial port.  This component converts the physical
measurements into a DSP-specific message. Figure
\ref{FIG_DSP_YSI_DATA} depicts the YSI Sonde's specific message format, 
which includes each measurement data point, described in Table
\ref{TAB_SONDE_MEASUREMENTS}.

@@

The \texttt{SondeDataContainerType} defines the payload of messages generated
by the YSI Sonde DP. As all DSP messages, it is derived from the
\texttt{AbstractMessageContent}. The payload mainly consists of the
\texttt{SondeDataType} that defines all the sensor data that the YSI Sonde is
capable of producing. The YSI Sonde DP uses the RS232 to query the
physical device for its data in regular intervals. The measurements
are then converted into a DSP message and forwarded to the DSP Broker,
from where they are forwarded to the remote communication component
described in the next section.  The YSI Sonde DP can also be
configured via special Update DSP messages, to direct the Sonde to change its
sampling frequency.  These property messages originate from a web
management interface (component \ding{197} in Figure \ref{FIG_NETBEAMS}).

@@

All messages sent by the YSI Sonde DP need to be uploaded to a remote
database. Since the DSP itself can only handle communication within
its local DPs and DCs, we have created a matching pair of DC and DP
that can handle remote communication (components \ding{193} and \ding{194} in Figure
\ref{FIG_NETBEAMS}). As explained earlier, we make use of the cellular
phone network to upload sensor data to a remote backend server. While
the cellular phone network has sufficient coverage for more
environmental sensor networks, it is not cost effective to keep a data
link permanently open. The remote communication components can be
configured to upload buffered sensor data only in certain time
intervals. Having this in mind, DSP Messages are transported via HTTP
\cite{RFC2068}. In this way, the matching DP/DC pair for remote communication
has the following properties:

\begin{itemize}
\item DSP Wire Transport Client: takes a collection of DSP messages
  and transmit them via HTTP. As a non-functional requirement, the
  interval in which this component will transmit messages to other
  components can be configured via the management interface;
\item DSP Wire Transport Server: is an HTTP server that receives the
  collection of DSP messages sent by the client, and piggy backs any
  messages destined for the remote client that may have queued up in
  the meantime.
\end{itemize}

The HTTP request and response are used to upload and download any
messages that have been queued at the respective end of the
communication link. The client and server components are capable of
buffering messages until they are sent to the remote destination
during the next communication cycle. Once messages have been
transported to their final destinations, they are unmarshalled and
passed on to the local DSP Broker for further dispatching. In this
regard, the client and server components are completely symmetrical. Messages
are dispatched with the help of the matcher to their final destinations.
